---
output: github_document
---

```{r}
#| label: setup-chunk
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# banrepstats

```{=html}
<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/banrepstats)](https://CRAN.R-project.org/package=banrepstats)
<!-- badges: end -->
```

`banrepstats` downloads economic time series from Banco de la República's
(Colombia's central bank) public statistics API — the same data behind its
[SUAMECA](https://suameca.banrep.gov.co/estadisticas-economicas/#/) portal —
and returns them as tidy `data.table` objects, ready to explore or plot.

## Installation

`banrepstats` is not on CRAN yet. Install it from GitHub with:

```r
# install.packages("remotes")
remotes::install_github("fersalme/banrepstats", build_vignettes = TRUE)
```

```{r}
#| label: setup
library(banrepstats)
```

## Browsing the indicator catalog

Banco de la República organizes its indicators in a menu tree (topic ->
sub-topic -> indicator). `catalogo_banrep()` downloads that tree from the
live API and flattens it into a single table, one row per indicator:

```{r}
#| label: catalogo
catalogo <- catalogo_banrep()
nrow(catalogo)
head(catalogo)
```

Each row is one indicator:

- `idGrupo` / `NombreGrupo`: the id and name of the parent topic (e.g.
  "Precios e inflación").
- `idSerie` / `NombreSerie`: the id and human-readable name of the
  indicator itself (e.g. "Inflación y meta").
- `idNombreSerie`: the indicator's unique code. This is the value you
  pass to `banrep_data()`.

Not sure which code you need? `search_banrep()` looks up a keyword
(case-insensitive, partial match) against the topic and indicator names:

```{r}
#| label: search-banrep
search_banrep("inflación")[, .(NombreGrupo, NombreSerie, idNombreSerie)]
```

## Downloading data

Pass one of those codes to `banrep_data()` to download the underlying
series as a single `data.table`:

```{r}
#| label: banrep-data-uno
precios <- banrep_data(indicator = "inflacion_y_meta")
head(precios)
```

A single indicator can still bundle more than one related series —
notice how `"inflacion_y_meta"` alone already returns two:

```{r}
#| label: banrep-data-series
unique(precios$id)
```

You can request several indicators at once — the results come back stacked
in a single `data.table`:

```{r}
#| label: banrep-data-varios
datos <- banrep_data(c("inflacion_y_meta", "tasas_interes_politica_monetaria"))
unique(datos$id)
```

Fetching the catalog on every call has a network cost. If you don't need
the live catalog, pass the bundled snapshot via `cache` to skip that
request:

```{r}
#| label: banrep-data-cache
precios_cache <- banrep_data("inflacion_y_meta", cache = banrep_cache)
identical(precios$value, precios_cache$value)
```

## Plotting a series

```{r}
#| label: plot-precios
#| fig-width: 7
#| fig-height: 4
series_plot(precios)
```

`series_plot()` draws one line per distinct series (the `id` column),
colored with the package's Banco de la República-inspired palette.

## What's included

- `catalogo_banrep()` — download and flatten the live indicator catalog.
- `search_banrep()` — search the catalog by keyword.
- `banrep_data()` — download one or more indicators as a single `data.table`.
- `series_plot()` — plot a `data.table` returned by `banrep_data()`.
- `banrep_cache` — a bundled snapshot of the catalog, for offline lookups.

See `vignette("banrepstats")` for a longer walkthrough, including an
example of integrating with the `tsibble`/`fable` forecasting ecosystem.

## License

MIT
