---
title: "banrepstats"
vignette: >
  %\VignetteIndexEntry{banrepstats}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

`banrepstats` downloads economic time series from Banco de la República's
(Colombia's central bank) public statistics API — the same data behind its
[SUAMECA](https://suameca.banrep.gov.co/estadisticas-economicas/#/) portal —
and returns them as tidy `data.table` objects, ready to explore or plot.

This vignette walks through the functions the package exports:

- `catalogo_banrep()`, to browse the indicator catalog,
- `search_banrep()`, to search the catalog by keyword,
- `banrep_data()`, to download one or more indicators, and
- `series_plot()`, to plot the result.

## Installation

`banrepstats` is not on CRAN yet. Install it from GitHub with:

```r
# install.packages("remotes")
remotes::install_github("fersalme/banrepstats")
```

```{r}
#| label: setup
library(banrepstats)
```

## Browsing the indicator catalog

Banco de la República organizes its indicators in a menu tree (topic ->
sub-topic -> indicator). `catalogo_banrep()` downloads that tree from the
live API and flattens it into a single table, one row per indicator:

```{r}
#| label: catalogo
catalogo <- catalogo_banrep()
nrow(catalogo)
head(catalogo)
```

Each row is one indicator:

- `idGrupo` / `NombreGrupo`: the id and name of the parent topic (e.g.
  "Precios e inflación").
- `idSerie` / `NombreSerie`: the id and human-readable name of the
  indicator itself (e.g. "Inflación y meta").
- `idNombreSerie`: the indicator's unique code. This is the value you
  pass to `banrep_data()`.

For example, "Inflación y meta" (under the "Precios e inflación" topic)
has the code `"inflacion_y_meta"`:

```{r}
#| label: catalogo-filtro
subset(catalogo, idNombreSerie == "inflacion_y_meta")
```

Not sure which code you need? `search_banrep()` looks up a keyword
(case-insensitive, partial match) against the topic and indicator names:

```{r}
#| label: search-banrep
search_banrep("inflación")[, .(NombreGrupo, NombreSerie, idNombreSerie)]
```

`search_banrep()` also accepts a `cache` argument, with the same meaning
as in `banrep_data()` (see below).

## Downloading data

Pass one of those codes to `banrep_data()` to download the underlying
series as a single `data.table`:

```{r}
#| label: banrep-data-uno
precios <- banrep_data(indicator = "inflacion_y_meta")
head(precios)
```

Each row is one observation of one series:

- `id`: the series name (e.g. "Meta de inflación").
- `unidad`: the unit of measurement (e.g. "Porcentaje").
- `periodicidad`: the reporting frequency (e.g. "Anual", "Mensual").
- `date` / `value`: the observation date and its value.

A single indicator can still bundle more than one related series —
notice how `"inflacion_y_meta"` alone already returns two:

```{r}
#| label: banrep-data-series
unique(precios$id)
```

### Multiple indicators at once

`indicator` accepts a vector of codes; the results come back stacked in a
single `data.table`:

```{r}
#| label: banrep-data-varios
datos <- banrep_data(c("inflacion_y_meta", "tasas_interes_politica_monetaria"))
unique(datos$id)
```

### Invalid indicator codes

Passing a code that isn't in the catalog raises an error listing the
valid options:

```{r}
#| label: banrep-data-error
#| error: true
#| eval: false
banrep_data("NO_EXISTE")
```

### Using the bundled catalog snapshot

`banrep_data()` calls `catalogo_banrep()` internally to validate the
indicator code and look up its id, which means every call hits the
catalog endpoint. If you're downloading many indicators, or don't need
the live catalog, fetch it once and pass it in via `cache`:

```{r}
#| label: banrep-data-cache
precios_cache <- banrep_data("inflacion_y_meta", cache = catalogo)
identical(precios$value, precios_cache$value)
```

The package also bundles a snapshot of the catalog as `banrep_cache`, so
you can look indicators up offline without calling `catalogo_banrep()` at
all:

```{r}
#| label: banrep-cache
head(banrep_cache)
```

## Plotting a series

`series_plot()` takes the `data.table` returned by `banrep_data()` and
draws one line per distinct series (the `id` column), colored with the
package's Banco de la República-inspired palette:

```{r}
#| label: plot-precios
#| fig-width: 7
#| fig-height: 4
series_plot(precios)
```

Series with different units are still placed on the same panel, so
`series_plot()` is best used with indicators that share a unit of
measurement:

```{r}
#| label: plot-tasas
#| fig-width: 7
#| fig-height: 4
tasas <- banrep_data("tasas_interes_politica_monetaria")
series_plot(tasas)
```

## Integrate with time-series forecasting packages

`banrep_data()` returns a plain `data.table`, not a `tsibble`, so it plugs
into the [tidyverts](https://tidyverts.org/) forecasting ecosystem
(`tsibble`, `fable`, `fable.prophet`) with one conversion step: turn the
`date` column into an explicit time index and declare the series with
`as_tsibble()`.

These packages are not dependencies of `banrepstats`, so this example is
not run when the vignette is built — install `tsibble`, `fable`,
`fable.prophet` and `feasts` yourself to try it. The plot below is a
pre-rendered snapshot of the code's output.

```{r}
#| eval: false
library(dplyr)
library(tsibble)
library(fable)
library(fable.prophet)
library(ggtime)

desempleo <- banrep_data(
  "mercado_laboral_poblacion_tasa_ocupacion_desempleo"
) |>
  filter(id == "Tasa de desempleo - total nacional") |>
  mutate(mes = yearmonth(date)) |>
  as_tsibble(index = mes)

# Hold out the last observations to check forecast accuracy later.
train <- desempleo |>
  filter(mes <= yearmonth("2024 Dec"))

fit <- train |>
  model(
    arima = ARIMA(value),
    ets = ETS(value),
    prophet = prophet(
      value ~ season(period = 12, order = 5, type = "multiplicative")
    )
  )
fit
#> # A mable: 1 x 3
#>                       arima          ets   prophet
#>                     <model>      <model>   <model>
#> 1 <ARIMA(0,1,0)(2,0,0)[12]> <ETS(M,N,A)> <prophet>

fc <- fit |> forecast(h = "18 months")
fc |> autoplot(desempleo)
```

![18-month forecast of Colombia's national unemployment rate (ARIMA, ETS and prophet)][fig-forecast]

[fig-forecast]: banrepstats-forecast.png